---
title: "Variation"
format:
  html: 
    toc: false
    section-divs: true
    include-after-body: 
      - text: |
          <script src="/js/progressive-reveal.js"></script>
          <script src="/js/bootstrapify.js"></script>
---

```{r include=FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 6 * 0.618,
  fig.retina = 3,
  dev = "ragg_png",
  fig.align = "center",
  out.width = "70%"
)

library(tidyverse)
library(checkdown)

source(here::here("R", "helpers.R"))
```


### ¿Qué es la variación? {.no-hide}

La **variación** es la tendencia de los valores a cambiar de medición en medición. Puedes ver variación muy fácilmente en la vida real: si mides cualquier variable continua dos veces, y de manera muy precisa, obtendrás dos resultados diferentes. Esto es cierto incluso si mides cantidades que son constantes, como la velocidad de la luz. Cada una de tus mediciones incluye una pequeña cantidad de error que varía de medición a medición. Las variables categóricas también pueden variar si las mides a lo largo de diferentes objetos (por ejemplo, el color de los ojos de diferentes personas) o en diferentes momentos (por ejemplo, los niveles de energía de un electrón en diferentes momentos).

Cada variable tiene sus propios patrones de variación, que pueden revelarnos informacion útil. La mejor manera de entender un patrón es visualizar la distribución de los valores de la variable. Cómo visualizar la distribución de una variable dependerá de si la variable es **categórica** o **continua**.

### Categorical variables

A variable is **categorical** if it can take only one of a small set of values. In R, categorical variables are usually saved as factors or character vectors. You can visualize the distribution of a categorical variable with a bar chart, like the one below. 

```{r echo=FALSE, out.width="80%"}
ggplot(data = mpg) +
  geom_bar(aes(x = class)) + 
  labs(title = "Classes of cars in mpg")
```

Don't worry if you cannot make or interpret a bar chart. We'll survey several types of charts in this tutorial, as we create a strategy for EDA. You'll learn how to build each type of chart in the tutorials that follow.

### Continuous variables

A variable is **continuous** if it can take any of an infinite set of smooth, ordered values. Here, smooth means that if you order the values on a line, an infinite number of values would exist between any two points on the line. For example, an infinite number of values exists between 0 and 1, e.g. 0.9, 0.99, 0.999, and so on. 

Numbers and date-times are two examples of continuous variables. You can visualize the distribution of a continuous variable with a histogram, like the one below:

```{r echo=FALSE, message=FALSE, out.width="80%"}
ggplot(data = faithful) +
  geom_histogram(aes(x = eruptions), binwidth = 0.25, color = "white", linewidth = 0.25) + 
  labs(title = "Lengths of 272 eruptions of Old Faithful")
```

### Frequencies

In both bar charts and histograms, tall bars show the common values of a variable, i.e. the values that appear frequently. Shorter bars show less-common values, i.e. values that appear infrequently. Places that do not have bars reveal values that were not seen in your data. To turn this information into useful questions, look for anything unexpected:

* Which values are the most common? Why?

* Which values are rare? Why? Does that match your expectations?

* Can you see any unusual patterns? What might explain them?

* Are there any **outliers**, which are points that don't fit the pattern or fall far away from the rest of the data? Are they the result of data entry errors or something else?

Many of the questions above will prompt you to explore a relationship *between* variables, to see if the values of one variable can explain the values of another variable. We'll get to that shortly.

### Review 4: Frequencies

The bar chart below visualizes the distribution of the `class` variable in the `mpg` data set, which comes in the {ggplot2} package. The height of the bars reveal how many cars in the data set come from each class. 

```{r echo=FALSE, out.width="80%"}
ggplot(data = mpg) +
  geom_bar(aes(x = class)) + 
  labs(title = "Classes of cars in mpg")
```

::: {.callout-note appearance="simple" icon=false .question}

**What is the most common type of car in the `mpg` dataset?**

```{r echo=FALSE}
check_question(
  answer = "suv", 
  options = c(
    "2seater",
    "compact",
    "midsize",
    "minivan",
    "pickup",
    "subcompact",
    "suv"
  ), 
  type = "radio", 
  button_label = "Registrar respuesta", 
  q_id = 3,
  right = c("Correct!")
)
```
:::

::: {.callout-note appearance="simple" icon=false .question}

**What is the least common type of car in the `mpg` dataset?**

```{r echo=FALSE}
check_question(
  answer = "2seater", 
  options = c(
    "2seater",
    "compact",
    "midsize",
    "minivan",
    "pickup",
    "subcompact",
    "suv"
  ), 
  type = "radio", 
  button_label = "Registrar respuesta", 
  q_id = 4,
  right = c("Correct!")
)
```
:::

###

Does the distribution of cars in the `mpg` dataset seem to reflect the distribution of cars that you see on the road? Would your answer shape how you use this data?

### Clusters

For continuous variables, clusters of similar values suggest that subgroups exist in your data. To understand the subgroups, ask:

* How are the observations within each cluster similar to each other?

* How are the observations in separate clusters different from each other?

* How can you explain or describe the clusters?

* Why might the appearance of clusters be misleading?


### Review 5: Clusters

The histogram below shows the distribution of the `eruptions` variable in the `faithful` data set, which comes with R. `eruptions` shows the lengths (in minutes) of 272 eruptions of the Old Faithful geyser in Yellowstone National Park.

To interpret the histogram, look first at the x axis, which displays the lengths of eruptions recorded in the data. The range of the x axis shows that the shortest eruptions lasted for about one minute and the longest for about five minutes. 

To see how many eruptions lasted for a specific length of time, find the length of time on the x axis and then look at the height of the bar above the length of time. For example, according to the histogram, 30 eruptions lasted for about two minutes, but only three lasted for about three minutes (the height of the bar above two is 30, the height of the bar above three is three).

```{r echo=FALSE, message=FALSE, out.width="80%"}
ggplot(data = faithful) +
  geom_histogram(aes(x = eruptions), binwidth = 0.25, color = "white", linewidth = 0.25) + 
  labs(title = "Lengths of 272 eruptions of Old Faithful")
```

::: {.callout-note appearance="simple" icon=false .question}

**Do the eruption lengths cluster into groups? How many?**

```{r echo=FALSE}
check_question(
  answer = "Correct thing", 
  options = c(
    "No. There are no clusters.",
    "Yes. Two clusters.",
    "Yes. Three clusters.",
    "Yes. Four clusters."
  ), 
  type = "radio", 
  button_label = "Registrar respuesta", 
  q_id = 5,
  right = c("Correct! Eruption lengths appear to be clustered into two groups: there are short eruptions (of around 2 minutes) and long eruptions (4-5 minutes), but few eruptions in between.")
)
```
:::


##

```{r}
#| echo: false
#| results: asis
create_buttons("03-covariation.html")
```
